---
title: "Case Study for Google Data Analytics Cyclistic"
author: "Francis"
date: "07/06/2021"
output:
  html_document: default
  pdf_document: default
---
# An analysis of the usage patterns of Casual users compared to Members of cyclistic bikes

In order to do this analysis, I used data available here [link](https://divvy-tripdata.s3.amazonaws.com/index.html)

This publicy available data was downloaded and stored as local files in a local folder on my pc.
The data is saved in quaters, and each quarter is a single csv file.  The last four quarters from Q2-2019 to q1-2020 was considered for the analysis since that made up the most recent full year (12 months) of data available.
```{r libraries used for the analysis}
#install.packages("lubridate")
#install.packages("ggplot2")
#install.packages("tidyverse")
```
```{r loading libraries}
library(lubridate)
library(ggplot2)
library(tidyverse)
```
```{r loading files}
getwd()
setwd("C:/Users/ekowh/OneDrive/Desktop/Documents/R School") 

rider_q1_20 <- read_csv("Divvy_Trips_2020_Q1.csv")
rider_q2_19 <- read_csv("Divvy_Trips_2019_Q2")
rider_q3_19 <- read_csv("Divvy_Trips_2019_Q3.csv")
rider_q4_19 <- read_csv("Divvy_Trips_2019_Q4.csv")

```
## inspecting the contents of each dataframe in order to match all fields and maintain a single name for all matching columns in each dataframe

```{r inspecting df 1}
head(rider_q1_20)
```

```{r inspecting df 2}
head(rider_q2_19)
```
```{r inspecting df 3}
head(rider_q3_19)
```

```{r inspecting df 4}
head(rider_q3_19)
```

```{r matching and renaming fields}
q1_2020<- rider_q1_20

q4_2019 <- rename(rider_q4_19
                  ,ride_id = trip_id
                  ,rideable_type = bikeid 
                  ,started_at = start_time  
                  ,ended_at = end_time  
                  ,start_station_name = from_station_name 
                  ,start_station_id = from_station_id 
                  ,end_station_name = to_station_name 
                  ,end_station_id = to_station_id 
                  ,member_casual = usertype)


q3_2019 <- rename(rider_q3_19
                  ,ride_id = trip_id
                  ,rideable_type = bikeid 
                  ,started_at = start_time  
                  ,ended_at = end_time  
                  ,start_station_name = from_station_name 
                  ,start_station_id = from_station_id 
                  ,end_station_name = to_station_name 
                  ,end_station_id = to_station_id 
                  ,member_casual = usertype)

q2_2019 <- rename(rider_q2_19
                  ,ride_id = "01 - Rental Details Rental ID"
                  ,rideable_type = "01 - Rental Details Bike ID" 
                  ,started_at = "01 - Rental Details Local Start Time"  
                  ,ended_at = "01 - Rental Details Local End Time"  
                  ,start_station_name = "03 - Rental Start Station Name" 
                  ,start_station_id = "03 - Rental Start Station ID"
                  ,end_station_name = "02 - Rental End Station Name" 
                  ,end_station_id = "02 - Rental End Station ID"
                  ,member_casual = "User Type")
```


## Inspecting the matched dataframes 
```{r inspecting the dataframes}
str(q1_2020)
str(q4_2019)
str(q3_2019)
str(q2_2019)
```
## Converting ride_id and rideable_type to chr

```{r convert columns to chr}
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
```

## Combining all separate dataframes into 1 big dataframe

```{r combine into 1 dataframe}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```
## Some columns were no more used in 2020 first quarter and these columns therefore have become inconsistent with the rest of the combined data and must be removed.

```{r removing inconsistent columns}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))
```

## Inspecting the new dataframe

```{r column names}
colnames(all_trips)
```


```{r number of rows}
nrow(all_trips)
```

```{r dim}
dim(all_trips)
```

```{r head}
head(all_trips)
```

```{r tail}
tail(all_trips)
```

```{r str}
str(all_trips)
```

```{r Summary}
summary(all_trips)
```

## In the head results we see that the member_casual field contains subscriber, and in the tail it contains member or casual, we need to inspect this column further.

```{r inspect member_casual column}
table(all_trips$member_casual)
```
## We notice that we have two values for casual i.e. subscriber/casual and two values for member i.e. Customer/member but all these only describe either a member or a casual customer, this column must therefore be updated to only contain member or casual.

```{r updating member_casual column}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual,"Subscriber" = "member","Customer" = "casual"))

```

```{r inspecting columns after update}
table(all_trips$member_casual)
```

## Breaking the date down into day, month, year and week for analysis

```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

## The most recent data, 2020 firts quarter did not have a ride duration column, therefore a ride duration must be calculated on across the entire dataframe for uniformity.

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

##Inspecting the dataframe

```{r}
str(all_trips)
```

## Converting ride_length to numeric for analysis
```{r}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```
## Viewing a summary of the ride_length just created
```{r}
summary(all_trips$ride_length)
```
Looks like there are certain rides with a negative length, this is an abnormality(bad data) and needs to be removed from the data in order to have objective analysis.

```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
summary(all_trips_v2$ride_length)
```

## Comparing members and casual trips

```{r average ride_length}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
```
```{r median}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
```


```{r max}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
```

```{r min}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

## Ride time by day of week between members and casual

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

## Average ride by month member_casual
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$month, FUN = mean)
```

## Average ride by Year member_casual
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$year, FUN = mean)
```

##Anaylsing data by weekday and member_casual
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  
  group_by(member_casual, weekday) %>%  
  summarise(number_of_rides = n()							 
            ,average_duration = mean(ride_length)) %>% 		
  arrange(member_casual, weekday)	
```
##Anaylsing data by Month and member_casual
```{r}
all_trips_v2 %>% 
  mutate(months = month(started_at, label = TRUE)) %>%  
  group_by(member_casual, months) %>%  
  summarise(number_of_rides = n()							 
            ,average_duration = mean(ride_length)) %>% 		
  arrange(member_casual, months)	
```
## Analysing data by year and member_casual
```{r}
all_trips_v2 %>% 
  mutate(r_year = year(started_at)) %>%  
  group_by(member_casual, r_year) %>%  
  summarise(number_of_rides = n()							 
            ,average_duration = mean(ride_length)) %>% 		
  arrange(member_casual, r_year)	
```

## Visualising Data, number of rides taken analysed by member_casual, weekday

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")+
labs(title="Number of Rides Taken",subtitle = "Analysed by Weekdays and Customer Type")
```
## Visualising Data, number of rides taken analysed by member_casual, month
```{r}
all_trips_v2 %>% 
  mutate(months = month(started_at, label = TRUE)) %>% 
  group_by(member_casual, months) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, months)  %>% 
  ggplot(aes(x = months, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(title="Number of Rides Taken",subtitle = "Analysed by Month and Customer Type")
```
## Visualising Data, number of rides taken analysed by member_casual, year

```{r}
all_trips_v2 %>% 
  mutate(year = year(started_at)) %>% 
  group_by(member_casual, year) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, year)  %>% 
  ggplot(aes(x = year, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(title="Number of Rides Taken",subtitle = "Analysed by Year and Customer Type")
```
## Visualising Data, average duration of rides analysed by member_casual, weekday

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(title="Average Duration of Rides",subtitle = "Analysed by Weekdays and Customer Type")

```

## Visualising Data, average duration of rides analysed by member_casual, month

```{r}
all_trips_v2 %>% 
  mutate(month = month(started_at, label = TRUE)) %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, month)  %>% 
  ggplot(aes(x = month, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(title="Average Duration of Rides",subtitle = "Analysed by Month and Customer Type")

```



## Visualising Data, average duration of rides analysed by member_casual, year

```{r}
all_trips_v2 %>% 
  mutate(weekday = year(started_at)) %>% 
  group_by(member_casual, year) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, year)  %>% 
  ggplot(aes(x = year, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(title="Average Duration of Rides",subtitle = "Analysed by Year and Customer Type")

```